---
title: "edaDataCamp"
author: "Brian Linn"
date: "February 20, 2017"
output: html_document
---
#Exploratory Data Analysis - Data Camp
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
remove(list = ls())
library(dplyr)
library(readr)
getwd()
comics <- read.csv("./data/comics.csv")
comics$gender <- comics$SEX
comics$align <- comics$ALIGN
comics$id <- comics$ID
glimpse(comics)

```
##Exploring Categorical Data
```{r}
# Print the first rows of the data
#comics

# Check levels of align
levels(comics$align)

# Check the levels of gender
levels(comics$gender)

# Create a 2-way contingency table
table(comics$align, comics$gender)
```

Removing the comics$align with fewest records
```{r}
#first print a count of the align variabls
table(comics$align)

# Load dplyr
library(dplyr)

# Remove align level
comics <- comics %>%
  filter(align != "Reformed Criminals") %>%
  droplevels()
```

###graphing alignment and gender to find view relationship
```{r}
# Load ggplot2
library(ggplot2)

# Create side-by-side barchart of gender by alignment
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "dodge")

# Create side-by-side barchart of alignment by gender
ggplot(comics, aes(x = gender, fill = align)) + 
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 90))
```

###counts and proportions
```{r}
options(scipen = 999, digits = 3) #simplify display format
comics$id <- comics$ID
#first create a table with the id type by alignment
tab_cnt <- table(comics$id, comics$align)
tab_cnt
#then use that table as input for the prop.table function to see the proportions in each category
#joint proportions
prop.table(tab_cnt)
#since the whole dataset was used, the total proportions = 1
sum(prop.table(tab_cnt))

#systematic relationship between variables
#conditional proportions - public/good
prop.table(tab_cnt, 1) #conditions on rows
prop.table(tab_cnt, 2) #conditions on columns

ggplot(comics, aes(x = id, fill = align)) +
        geom_bar(position = "fill") +
        ylab("proportion")
#since the data is conditioned on the id - columns - use "fill" as geom position to make them add to 1
#now condition on alignment by swithching the x and fill in the aes argument
ggplot(comics, aes(x = align, fill = id)) +
        geom_bar(position = "fill") +
        ylab("proportion")


# Plot of gender by align - counts
ggplot(comics, aes(x = align, fill = gender)) +
  geom_bar()
  
# Plot proportion of gender, conditional on align - proportions
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "fill")
```

###distribution of one variable
```{r}
table(comics$id)
tab_cnt <- table(comics$id, comics$align)
tab_cnt

#univariate bar chart
ggplot(comics, aes(x = id)) + 
        geom_bar() +
        theme(axis.text.x = element_text(angle = 90))

#faceting
#looking at the neutral characters
ggplot(comics, aes(x = id)) + 
        geom_bar() +
        facet_wrap(~align) +
        theme(axis.text.x = element_text(angle = 90))

#pie charts - meh

#set neutral to the middle, and then plot align
# Change the order of the levels in align
comics$align <- factor(comics$align, 
                       levels = c("Unknown", "Bad Characters", "Neutral Characters", "Good Characters"))
levels(comics$align)
# Create plot of align
ggplot(comics, aes(x = align)) + 
  geom_bar()


# Plot of alignment broken down (faceted) by gender
ggplot(comics, aes(x = align)) + 
  geom_bar() +
  facet_wrap(~ gender)
```

#Exploring Numerical Data

```{r}
library(MASS)
data("Cars93")

cars <- Cars93
str(cars)

#dotplots
ggplot(cars, aes(x = Weight)) +
        geom_dotplot(dotsize = 0.4)

#histogram aggregates the dots into bins on the x-axis
ggplot(cars, aes(x = Weight)) +
        geom_histogram()

#density plot represents the shape of the histogram as a line
ggplot(cars, aes(x = Weight)) +
        geom_density()

#boxplot
ggplot(cars, aes(x = 1, y = Weight)) +
        geom_boxplot() +
        coord_flip()

#faceted histogram
ggplot(cars, aes(x = MPG.highway)) +
        geom_histogram() +
        facet_wrap(~Type)

# Load package
library(ggplot2)

# Learn data structure

cars$van <- ifelse(cars$Type == "Van", "TRUE", "FALSE")
cars$rear_wheel <- ifelse(cars$DriveTrain == "Rear", "TRUE", "FALSE")
cars$city_mpg <- cars$MPG.city
# Create faceted histogram
ggplot(cars, aes(x = city_mpg)) +
  geom_histogram() +
  facet_grid(. ~ van)

#using the DataCamp dataset
# Filter cars with 4, 6, 8 cylinders
cars$ncyl <- cars$Cylinders
cars$city_mpg <- cars$MPG.city
common_cyl <- filter(cars, ncyl %in% c(4, 6, 8))

# Create box plots of city mpg by ncyl
ggplot(common_cyl, aes(x = as.factor(ncyl), y = city_mpg)) +
  geom_boxplot()

# Create overlaid density plots for same data
ggplot(common_cyl, aes(x = city_mpg, fill = as.factor(ncyl))) +
  geom_density(alpha = .3)
```

###distribution of one variable
```{r}
ggplot(cars, aes(x = city_mpg)) +
        geom_histogram() +
        facet_wrap(~Type)
#buid the data pipeline - cannot facet for engine size since it is continuous
cars$eng_size <- cars$EngineSize
cars2 <- cars %>% 
        filter(cars$eng_size < 2.0)

ggplot(cars2, aes(x = city_mpg)) +
        geom_histogram() +
        facet_wrap(~Type)

#now combine into one data pipeline
cars$hwy_mpg <- cars$MPG.highway
cars %>% 
        filter(eng_size < 2.0) %>% 
        ggplot(aes(x = hwy_mpg)) +
        geom_histogram()
        
#bin widths - less bins tends to provide a smoother view of the data
cars %>% 
        filter(eng_size < 2.0) %>% 
        ggplot(aes(x = hwy_mpg)) +
        geom_histogram(binwidth = 5)

#same for density plots - except use bw for bandwidth
cars %>% 
        filter(eng_size < 2.0) %>% 
        ggplot(aes(x = hwy_mpg)) +
        geom_density(bw = 5)

#marginal and conditonal histograms
cars$horsepwr <- cars$Horsepower
cars$msrp <- cars$Price
#marginal distribution of horsepower
# Create hist of horsepwr
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  ggtitle("horsepower")

#conditional histogram of horsepower for affordable cars
# Create hist of horsepwr for affordable cars
cars %>% 
  filter(msrp < 25000) %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  xlim(c(90, 550)) +
  ggtitle("horsepower and affordability")

#different binwidths
# Create hist of horsepwr with binwidth of 3
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram(binwidth = 3) +
  ggtitle("horsepower")

# Create hist of horsepwr with binwidth of 30
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram(binwidth = 30) +
  ggtitle("horsepower")

# Create hist of horsepwr with binwidth of 60
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram(binwidth = 60) +
  ggtitle("horsepower")


```

###box plots
```{r}
# Construct box plot of msrp
cars %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()

# Exclude outliers from data
cars_no_out <- cars %>%
  filter(msrp < 100000)

# Create plot of city_mpg
cars %>%
  ggplot(aes(x = 1, y = city_mpg)) +
  geom_boxplot()

# Create plot of width
cars %>% 
  ggplot(aes(x = Width)) +
  geom_density()
#density works better as the second mode is hidden by the box plot
```

###visualizations in higher dimensions
```{r}
cars$van <- ifelse(cars$Type == "Van", "TRUE", "FALSE")
cars$rear_wheel <- ifelse(cars$DriveTrain == "Rear", "TRUE", "FALSE")
ggplot(cars, aes(x = msrp)) +
        geom_density() +
        facet_grid(van ~ rear_wheel, labeller = label_both)
#before tilde are the rows of the graphs, and after tilde are the columns
table(cars$rear_wheel, cars$van)

#3 variable plot
common_cyl <- filter(cars, ncyl %in% c(4, 6, 8))
common_cyl %>%
  ggplot(aes(x = hwy_mpg)) +
  geom_histogram() +
  facet_grid(ncyl ~ van) +
  ggtitle("mileage by cylinders for vans")
```

###measures of center

```{r}
library(ggplot2)
library(dplyr)
life <- read.csv("./data/dhsData/SUMMARYMEASURESOFHEALTH.csv")
glimpse(life)
life$expectancy <- life$ALE
x <- head(round(life$expectancy), 11)

xDf <- as.data.frame(x)

mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

ggplot(xDf, aes(x = x)) + 
        geom_dotplot() +
        geom_vline(xintercept = mean(x), colour = "red") +
        geom_vline(xintercept = median(x), colour = "blue") +
        geom_vline(xintercept = mode(x), colour = "gold")
#mean - average value
sum(x)/length(x)
mean(x)

#median - center value
sort(x)

#mode - most frequent occurence
table(x)

mode(x)

#Compare life expectancies of west coast to the rest of the US
life$state <- life$CHSI_State_Name
life <- life %>% 
        mutate(west_coast = state %in% c("California", "Oregon", "Washington"))
glimpse(life)

#creating groupwise means
life %>%
        group_by(west_coast) %>% 
        summarize(mean(expectancy),
                  median(expectancy))

#group by and summarize are a powerful combination of functions 
life %>% 
        slice(240:247) %>% 
        group_by(west_coast) %>% 
        summarize(mean(expectancy))

#class exercises
library(gapminder)
data(gapminder)
life2 <- gapminder

# Create dataset of 2007 data
gap2007 <- filter(gapminder, year == 2007)

# Compute groupwise mean and median lifeExp
gap2007 %>%
  group_by(continent) %>%
  summarize(mean(lifeExp),
            median(lifeExp))

# Generate box plots of lifeExp for each continent
gap2007 %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot()

sum(x - mean(x))
#result is close to 0, why didn't measure detect it - the negatives
#to calculate sample variance
(sum((x - mean(x))^2)) / (length(x) - 1)
var(x)

#sample standard deviation
sqrt(var(x))
sd(x)
#sd is 1.47 years
#variance is in years sqaured which has no intuitive reference

#inter-quartile range cuts off 50% of the data
#height of boxplot
summary(x)
IQR(x)

#range of x
range(x)
diff(range(x))

#which are most meaningful
#start with sd - but careful as this sensitive to outliers
#IQR is less sensitive to the outliers or skewed data

#class exercises
gap2007 <- gapminder %>% 
        filter(year == 2007)
# Compute groupwise measures of spread
gap2007 %>%
  group_by(continent) %>%
  summarize(sd(lifeExp),
            IQR(lifeExp),
            n())

# Generate overlaid density plots
gap2007 %>%
  ggplot(aes(x = lifeExp, fill = continent)) +
  geom_density(alpha = 0.3)

# Compute stats for lifeExp in Americas
gap2007 %>%
  filter(continent == "Americas") %>%
  summarize(mean(lifeExp),
            sd(lifeExp))

# Compute stats for population
gap2007 %>%
  summarize(median(pop),
            IQR(pop))

#shape of the distribution
#modality - the number of peaks in the data
#unimodal is the bell curve
#bimodal has two peaks
#multimodal has 3 or more
#uniform has no mode

#skewness
#right tail = right skewed and vice versa
#versus symmetric

#income west coast v rest of country
#need to add some income to life - lol
life <- read.csv("./data/dhsData/SUMMARYMEASURESOFHEALTH.csv")
life$expectancy <- life$ALE
life$state <- life$CHSI_State_Name
life <- life %>% 
        mutate(west_coast = state %in% c("California", "Oregon", "Washington"))

income <- read.csv("./data/dhsData/est15All.csv")
income$income <- as.numeric(levels(income$Median_Household_Income)[income$Median_Household_Income])

glimpse(income)

life <- merge(life, income[, c(1, 2, 6)], by = c("County_FIPS_Code", "State_FIPS_Code"))

glimpse(life)

ggplot(life, aes(x = income, fill = west_coast)) +
        geom_density(alpha = 0.3)

ggplot(life, aes(x = income)) +
        geom_histogram(binwidth = 60)

#use the log function to pull in the right tail skewness
ggplot(life, aes(x = log(income), fill = west_coast)) +
        geom_density(alpha = 0.3)

#class exercises with gapminder
library(gapminder)
data("gapminder")
gap2007 <- gapminder %>% 
        filter(year == 2007)

# Create density plot of population
gap2007 %>%
  ggplot(aes(x = pop)) +
  geom_density(alpha = 0.3)

# Transform the skewed pop variable
gap2007 <- gap2007 %>%
  mutate(log_pop = log(pop))

# Create density plot of new variable
gap2007 %>%
  ggplot(aes(x = log_pop)) +
  geom_density(alpha = 0.3)

#outliers
#center, variability, and shape
#outliers are far from the bulk of the disutribution
par(mfrow = c(1, 2))
ggplot(life, aes(x = income, fill = west_coast)) +
        geom_density(alpha = 0.3)

ggplot(life, aes(x = income, y = west_coast)) +
        geom_boxplot()+
        coord_flip()

#create a new column to classify each county as outlier or not
life <- life %>%
        mutate(is_outlier = income > 75000)

#view a df with the outliers
life %>% 
        filter(is_outlier) %>% 
        arrange(desc(income))

life %>% 
        filter(!is_outlier) %>% 
        ggplot(aes(x = income, fill = west_coast)) +
        geom_density(alpha = 0.3)

#plotting without the outliers
# Filter for Asia, add column indicating outliers
gap_asia <- gap2007 %>%
  filter(continent == "Asia") %>%
  mutate(is_outlier = lifeExp < 50)

# Remove outliers, create box plot of lifeExp
gap_asia %>%
  filter(!is_outlier) %>%
  ggplot(aes(x = 1, y = lifeExp)) +
  geom_boxplot()
```


###Case Study
Email data
```{r}
# Load packages
library(ggplot2)
library(dplyr)
library(openintro)

data(email)

# Compute summary statistics
email %>%
  group_by(spam) %>%
  summarize(median(num_char),
  IQR(num_char))

# Create plot
email %>%
  mutate(log_num_char = log(num_char)) %>%
  ggplot(aes(x = spam, y = log_num_char)) +
  geom_boxplot()

#write the below code from scratch
# Compute center and spread for exclaim_mess by spam
email %>%
  group_by(spam) %>%
  summarize(median(exclaim_mess),
  IQR(exclaim_mess))


# Create plot for spam and exclaim_mess
ggplot(email, aes(x = spam, y = log(exclaim_mess))) +
geom_boxplot()

ggplot(email, aes(x = log(exclaim_mess))) +
geom_histogram(bins = 60) +
facet_wrap(~spam)

ggplot(email, aes(x = log(exclaim_mess), fill = spam)) +
geom_density(aplha = 0.3)


#zero-inflation strategies
#analyze them separately or treat as categorical (zeroes and non-zeroes)

email$spamC <- ifelse(email$spam == 0, "not spam", "spam")
#side by side histogram
email %>% 
        mutate(zero = exclaim_mess == 0) %>% 
        ggplot(aes(x = zero)) +
        geom_bar() +
        facet_wrap(~spamC)
#barchart
email %>% 
        mutate(zero = exclaim_mess == 0) %>% 
        ggplot(aes(x = zero, fill = spamC)) +
        geom_bar()

#counts v proportions
#to get proportions add position = "fill to bar geom"
email %>% 
        mutate(zero = exclaim_mess == 0) %>% 
        ggplot(aes(x = zero, fill = spamC)) +
        geom_bar(position = "fill")
#results in conditional proportions

# Create plot of proportion of spam by image
email %>%
  mutate(has_image = image > 0) %>%
  ggplot(aes(x = has_image, fill = spam)) +
  geom_bar(position = "fill")

# Test if images count as attachments
sum(email$image > email$attach)

# Question 1
email %>%
  filter(dollar > 0) %>%
  group_by(spam) %>%
  summarize(median(dollar))

# Question 2
email %>%
  filter(dollar > 10) %>%
  ggplot(aes(x = spam)) +
  geom_bar()

#considering proportions and implications
email %>% 
        mutate(has_image = image > 0) %>% 
        ggplot(aes(x = as.factor(has_image), fill = spamC)) +
        geom_bar(position = "fill")
#proportion of spam to not spam conditioned on has_image

email %>% 
        mutate(has_image = image > 0) %>% 
        ggplot(aes(x = spam, fill = has_image)) +
        geom_bar(position = "fill")
#proportion of has image conditoned on spam

#ordering bars - not directly from code but from how r treats the variable
email <- email %>% 
        mutate(zero = exclaim_mess == 0)
levels(email$zero)
#null because it is a logical vector not a factor
#so we must factor it
email$zero <- factor(email$zero,
                     levels = c("TRUE", "FALSE"))
email %>% 
        ggplot(aes(x = zero)) +
        geom_bar() +
        facet_wrap(~spamC)

#class exercises
# Reorder levels
email$number <- factor(email$number,
                        levels = c("none", "small", "big"))

# Construct plot of number
email %>%
  ggplot(aes(x = number)) +
  geom_bar() +
  facet_wrap(~spam)


```
