---
title: "corrRegressDataCamp"
author: "Brian Linn"
date: "February 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DAAG)
library(dplyr)
library(ggplot2)
library(openintro)
```

### Bivariate Relationships
-Both variables are numerical 
        -output is the response - y - dependent
        -explanatory variables - inputs - x - independent or predictor         variables
        
        Response goes on y - vertical
        Explanatory goes on x - horizontal

ggplot

```{r}
data(possum)
glimpse(possum)
ggplot(data = possum, aes(y = totlngth, x = taill)) +
        geom_point() +
        scale_x_continuous("Length of Possum Tail (cm)") +
        scale_y_continuous("Length of Possum Body (cm)")

```

Can think of boxplots as scatterpots
        -with discretized explanatory variable

```{r}
#use cuts to break up the tail lengths into groups
possum %>% ggplot(aes(y = totlngth, 
                      x = cut(taill, breaks = 5))) +
        geom_boxplot() +
        scale_x_discrete("Length of Possum Tail (cm)") +
        scale_y_continuous("Length of Possum Body (cm)")

#class exercises with scatterplots
data("ncbirths")
glimpse(ncbirths)

# Scatterplot of weight vs. weeks
ggplot(data = ncbirths, aes(y = weight, x = weeks)) + 
        geom_point()

#boxplots as discretized scatterplots
#variable on x-axis is discretized - with cut and #of breaks
ncbirths %>% ggplot(aes(y = weight,
                        x = cut(weeks, breaks = 5))) +
        geom_boxplot()

        
```

###Characterizing bivariate relationships
look for
        -form (linear, quadratic, non-linear)
        -direction (positive or negative)
        -strength (how much scatter/noise?)
        -outliers
        
```{r}
#creating scatterplots
data("mammals")
data("mlbBat10")
data("bdims")
data("smoking")
glimpse(mammals)

#with mammals - brainweight by body weight
mammals %>% ggplot(aes(x = BodyWt, y = BrainWt)) +
        geom_point()

mammals %>% ggplot(aes(x = log(BodyWt), y = log(BrainWt))) +
        geom_point()

#baseball sluggin v on base percentage
glimpse(mlbBat10)
mlbBat10 %>% ggplot(aes(x = OBP, y = SLG)) +
        geom_point()

#bdims weight as function of height
glimpse(bdims)
bdims %>% ggplot(aes(x = hgt, y = wgt, color = factor(sex)))+
        geom_point()

#smoking amount smoked as predicted by age
glimpse(smoking)
smoking %>% ggplot(aes(x = age, 
                       y = (amtWeekends + amtWeekdays))) +
        geom_point()

#characterizing scatterplots
linear - negative - moderatley strong

```

```{r}
#transformations
mammals %>% ggplot(aes(x = BodyWt, y = BrainWt)) +
        geom_point() +
        scale_x_log10() +
        scale_y_log10()

mammals %>% ggplot(aes(x = BodyWt, y = BrainWt)) +
        geom_point() +
        coord_trans(x = "log10", y = "log10")
```

###Outliers
observations that don't fit with the rest of the points

```{r}
mlbBat10 %>% ggplot(aes(x = SB, y = HR)) +
        geom_point(alpha = 0.5, position = "jitter")

#indentify the outliers
mlbBat10 %>% 
        filter(SB>60 | HR > 50) %>% 
        select(name, team, position, SB, HR)
#slg by OBP with at bats greater than or equal to 200
mlbBat10 %>%
        filter(AB > 199) %>% 
        ggplot(aes(x = OBP, y = SLG)) +
        geom_point()

#find row with player(s) that have 200 at bats and OBP less than 0.2
mlbBat10 %>% 
        filter(AB > 199, OBP < 0.200) %>% 
        select(name, team, position, AB, OBP)

```

